Solution of Differential Algebraic Equations with Manifold Constraints on the Latent Space of an Autoencoder

### 2019-06-15 paper/paper.tex
equation for hessian augmentation
added a few citations to the paper

The system of equations we are looking at are, for only two components,
\begin{align}
  \partial_t \rho_{H_2O} =& \nabla \cdot F_{H_2O} + s \\
  \partial_t \rho_{CO_2} =& \nabla \cdot F_{CO_2} + s \\
  \partial_t e =& \nabla \cdot F_{heat} + s
\end{align}
More species add more balance laws (rows of equations). The applications also involve continuum
mechanics and chemical reactions, which are not considered. The interplay of deformations,
reactions, and phase transitions makes the equations extremely difficult
to solve. \cite{queiruga2018millstone}
(In this paper, we use use $\partial_t$, $\mathrm{d}/\mathrm{d}t$,
and $\dot{()}$ to denote differentiation with respect to time
interchangably depending on context.
The complixity is hidden inside the relationships between temperature
$T$, and pressure,$p$, density, $\rho$, and internal energy $e$ (or
enthalpy, $h$, s.t. $e=\rho h - p$), 
for the materials in question.

% {\bf RESERVOIR PHYSICS LIT REVIEW HERE; Other applications;
%   plasticity, }

@@

We redescribe our problem as a differential algebraic equation
with a constraint: the balances of mass and energy are constrained by
the EOS. We can cast the above equation into a basic form for a
differential algebraic equation,
\begin{align}
  z( u, \partial_t u) &= 0 \\
  c(u) & = 0
\end{align}
where, for the equation above, the variable is $u=\{T,p,\rho,h\}$. We will introduce a new way of thinking about constraints as sets of
solutions or observations of possible $u$:
\begin{equation}
c(u)=0 \longleftrightarrow \left\{ u | c(u)=0 \;\text{or}\; u\; \text{observed} \right\}
\end{equation}
This maps directly to the original experimental procedure for
determining a constraint equation.

@@

 A program synthesis approach was used to search for and evaluate possible methods of Peridynamics in Queiruga, 2017 \cite{queiruga2017numerical}. (An open source implementation of the study can be found Periflakes)\cite{queiruga2018periflakes}.

@@

We instead replace our problem with two balance laws and one unknown
constraint:
\begin{align}
\text{Solve for}\, \rho(t), \, h(t), \, p(t),\, \text{and}\, T(t)\, \text{satisfying:}\\
\partial_t \rho &  = \mathbf{k}_p(p_\infty - p) + r\\
\partial_t (\rho h-p) & = \mathbf{k}_T(T_\infty-T) + s \\
\text{such that they lie on the material EOS,}\\
\rho,h,p,T & \in \{\text{EOS}\}
\end{align}
The \emph{ground truth} is the experimental data in the first place; the
branching curve fit is only one realization of representing the data. We
have demonstrated that it is not possible to represent any two of these
in terms of the other two; and even if so, it is not necessarily a good
and robust fit. What if we search for a new fit?

@@

@@

The architecture, illustrated in Figure \ref{fig:classifyingnetwork}, has the form:
\begin{align}
  q &= W_{enc} x^n + b_{enc}\\
  f &= W_{dec} q^m + b_{dec} \\
  h &= \sigma\left(W_{bound}q^m+b_{bound}\right) \\
  p &= \text{softmax}\left(W_{phase}q^m+b_{bound[phase}\right) \\
  \bar{x}&= f * p
\end{align}

@@

\begin{figure}
  \centering
  \includegraphics[width=3in]{/Users/afq/Documents/Dropbox/ML/primaryautoencoder/slides/classifier_good.jpg}
  \caption{\label{fig:classified}Result of the unsupervised phase classifier. Compare this plot to the original labeled dataset Figure \ref{fig:pTrho}.}
\end{figure}
Because the identity of the phases,
For comparison purposes, we group together liquid-gas-supercritical as one contiguous label.

@@

With the trained model, we solve for \(q_1(t)\) and \(q_2(t)\) such that:
\begin{align}
\partial_t \rho(q_1,q_2) & = \nabla \cdot \mathbf{k} \nabla p(q_1,q_2) + r \\
\partial_t \rho h(q_1,q_2)-p & = \nabla \cdot \mathbf{k'}\nabla T(q_1,q_2) + s
\end{align}
where \(\rho(q_1,q_2)\) etc. are the back of an autoencoder:

Using the autoencoder to automatically remove the local constraint:

\begin{equation}
eos(D(q_1,q_2)) = 0 \, \forall \,q_1,q_2
\end{equation}

Only training the material representation, not the balance
laws.

First, we had a constrained DAE:
\begin{align}
\frac{\mathrm{d}}{\mathrm{d}t} m(T,p,\rho,h) &= r(T,p,\rho,h)\\
eos(T,p,\rho,h) &= 0
\end{align}
where $m=\{\rho,\rho h -p\}$ and $r$ is the right hand side, with no good representation of \(eos\) everywhere.

We obtain four new functions with two unknowns. We're actually solving the manifold of Figure X.
\begin{figure}
  \centering
  \includegraphics{../slides/water_four_surfaces_plot.png}
  \caption{\label{fig:foursurfaces}The four 2D surfaces that make up the new parameterization. The learned surface captures the sharp kinks and vertical fronts in Figure \ref{fig:classifyingresult}, but is ananalytic and smooth function of the two learned unknowns.}
  \end{figure}

The decoder part of the network is inserted into the balance law

\begin{equation}
\frac{\mathrm{d}}{\mathrm{d}t} m(D(q)) = r(D(q))
\end{equation}

We can parameterize this equation with any time stepping scheme, .e.g. a
Runge-Kutta scheme. For subsurface flow, backward Euler is popular due
to its simplicity and L-stability, which we follow since the topic of
this paper is not exploring new time stepping schemes. This yields the
discretizated algebraic equation for \(q_i\) and \(t+\Delta t\) given the
value \(q^0\) at \(t\),

@@

This equation is deceptively simple: the function \(R(q)\) has complex logic encoded in it; if-then
statements appear if the model used activations such as rectifiers.
However, the complex material behavior which traditionally requires
the vast majority of programming was implemented automatically by the
models; implementing equations XX, XX, and XX requires only a few lines
of human-written code.

@@

There is a caveat with representing the degrees of freedom of the system
on a learned space: the user does not know the meaning of \(q\) as an
input. Further, the values of \(q\) vary from network to network and
depend on the parameterization. The user needs to be able to specify the
"real" physical quantities.

Since the original manifold is 2D, the user only needs to specify two
initial values. In a multiphase solver such as TOUGH, the state needs
to be specified with the two primary variables plus the state tag (sans input file formatting): 
\begin{verbatim}
    9e5 20 Aqu
    0.5 20 AqG
\end{verbatim}
Note that the first primary variable means pressure in the first line
for a pure-liquid state,
but saturation in the second line for the liquid-gas equilibrium.
This makes sense inuitively, as this is how humans are taught, but
still leads to bugs and occasionally requires looking up values in
textbooks or the manual. (Is it $S_{gas}$ or $S_{aqu}$ on the second
line?) A fully fledged solver needs to check the user specified
initial conditions for erroneous inputs.

@@

Note that the encoder only necessarily works when all four values are
correct. Thus, it is necessary to actually solve the nonlinear problem
on the decoder:
\[D(q^0)_i = s^0_i \quad \text{for specified}\; i\]
This equation has two rows indexing into the four state variables;
e.g. only the rows associated with $p$ and $T$ are solved if those are
specified, which is usually the case. The above equation is solved
with Newton's method. The two corresponding rows are cut out of
the tangent matrix to $D$,
\[K_{ij} = \frac{\partial D_i}{\partial q_j}  \quad \text{for specified}\; i, \text{for}\;j=1,2\]
and the following system of equations is solved for an update to $q$,
\[K_{ij}\Delta q_j = s^0_i-D(q)_i\]
The iteration u $q$ is repeated until convergence, with under-relaxation used again to update
 \begin{equation}
q^0_{k+1} =q^0_{k} + \alpha \Delta q_{k}
\end{equation}
with the under-relaxation parameter usually set to $\alpha \approx
0.1$.
The initial condition problem is not gaurunteed to converge, however,
and often does not.
For example, if the initial guess was in the liquid phase and the target point
in the EOS space is in a solid, the iteration will fail.
To address this issue, we perform multiple starts with an initial condition in each
phase and equilibrium, discard points that diverge, and then pick the
\(q\) that is closest to the two specified components. The multiple
start points are specified at EOS development time to form a set of
values that represent middle points for each phase,
\begin{equation}
\left\{s_{guess}\right\} = \left\{ \left(\begin{array}{c}
\bar{T}_{gas}\\
\bar{p}_{gas}\\
\bar{\rho}_{gas}\\
\bar{h}_{gas}
\end{array}\right),\left(\begin{array}{c}
\bar{T}_{liq}\\
\bar{p}_{liq}\\
\bar{\rho}_{liq}\\
\bar{h}_{liq}
\end{array}\right),\left(\begin{array}{c}
\bar{T}_{ice}\\
\bar{p}_{ice}\\
\bar{\rho}_{ice}\\
\bar{h}_{ice}
\end{array}\right)... \right\}
  \end{equation}
The multistart method could be specified without any prior knowledge
with random points, but this is a sufficiently robust method for
now. Each of the possible guess points is used to fill in the two
unspecified values as guesses and then passed into the encoder.
E.g. if $p$ and $T$ were specified, each possible $\bar{\rho}$ and
$\bar{h}$ combination from the above sets are filled in to provide
multiple guesses for $q$,
\begin{equation}
s^0 = \left(\begin{array}{c}
T^*(0)\\
p^*(0)\\
\bar{\rho}_{guess}\\
\bar{h}_{guess}
\end{array}\right);\quad q^0_{guess}=E\left(s^0\right)
 \end{equation}
 These sets of $q^0_{guess}$ are then iterated upon using the Newton's
 method procedure.

 It is further possible that none of the initial guesses for this iteration
converge. There are two possiblities: the target state point is
incorrectly specified or out-of-range (e.g., negative density), or the
network is incapable of representing the desired material properties
(i.e., poorly trained).

The weakness of this necessity means that more error checking is needed
in verifying user inputs, but this is necessary in any
multiphase/multicomponent reservoir simulator. LatentSim does not yet
detect and report errors, and only reports divergence. In our study, the
testing problem specifications are all correct, but not all network
realizations are acceptable. Failure to determine a \(q\) for one of the
test initial conditions is a common mode of detecting unacceptable
network realizations; this is marked as a testing error during training
which invalidates a model as a potential realization.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Make a database of \(T,p,\rho,h,X_1,...\)

  \begin{itemize}
  \item
    Piece together empirical fits for each phase from literature
  \item
    (\emph{Experimental data in the future})
  \end{itemize}
\item
  Normalize (and \(\log(p)\)) the database, shuffle it

@@

The codebase is released open source at
\url{https://github.com/afqueiruga/latentsim}. Helper methods that the
author reuses in various TensorFlow-based projects are factored out in
the dully-named \url{https://github.com/afqueiruga/afqstensorflowutils}.
The implementation of the empirical equations of state for water are
factored out at
\href{https://github.com/afqueiruga/equations_of_state}{https://github.com/afqueiruga/equations\_of\_state}.

@@

Three different extents:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Linear Equation of State

  \begin{itemize}
  \item
    \textbf{Test!}
  \item
    Reduces to single phase Darcy's law problem (slide 1)
  \item
    \( p = 10^5+[-10^3, 10^3] Pa,\quad T = [ 19, 21 ] ^o C\)
  \end{itemize}
\item
  Water Liquid-Gas Regime

  \begin{itemize}
  \item
    One phase boundary 
  \item
    \( p = [100,5\times 10^5] Pa, \quad T = [274,594] K\)
  \end{itemize}
\item
  Water Solid-Liquid-Gas-Supercritical Regimes

  \begin{itemize}
  \item
    No linear mapping to latent space 
  \item
    Entire span
  \item
    \( p = [6\times 10^{-6},3\times 10^8]Pa, \quad T = [150,1000] K \)
  \end{itemize}
\end{enumerate}

@@

The verification problems are listed in Table \ref{tab:problems}.
\begin{table}
  \caption{\label{tag:problems}Enumeration of the problems used to test and evaluate each fo the generaed architectures}
\begin{tabular}{l|c|c|c}
  Name & start & path & end\\
\hline
\end{tabular}
\end{table}
The simulator was benchmarke against the long standing TOUGH+ multiphase
multicomponent

Single-grid block simulations were set up for most of the cases. TOUGH+
does not handle supercritical \(H_2O\), so these test problems were not
used to benchmark. (The supercritical phase region is not practically
relevant for water, but it is very important for other fluids such as
\(CO_2\).) (Note that, as a major motivation of this work, adding
supercritical states to the existing Fortran simulator would require
coding of one new logical state with two new state transitions with
primary variable switches; for the approach of this paper, the dataset
just needed to be extended.)

@@

We illustrated a programming model that is able to incorporate machine
learning models into existing well known computational methodologies.
This allows us to enforce physically known quanities, such as the
balance laws.

The TensorFlow system is not quite suited for this type of workflow -\/-
the resulting simualtions are extremely slow due to repeated calls to
session.run and a very complicated graph structure for the polynomials.
Currently, the system is being reimplemented in Julia using Flux and
Zygote to seemlessly reincorporate into a fully-fledged new reservoir
simulator. The pendulum has been reimplemented in Julia at .

@@

Each implemented equation of state (three options in this work spanning
different extents of \(H_2O\)) has a database of potential realizations.
The entries are model architectures, parameters, and training
environment configurations and logs for each equation-of-state dataset.
This database of models replaces a version controlled repository of user
written modules. Hundreds of trials for novel primary variable
configurations happen in a matter of minutes on a workstation in place
of pain-staking trial-and-error labor performed with human "intuition"
and experience. We did encode a little bit of human intuition into the
model architectures and tricks for initializing the networks.

@@

In TOUGH, for one instance, all of the calculations are performmed in SI units because it is simply too much labor to normalize everything! Adverse effects to iterative algorithms and floating point error propogation occur, especially because numerical differentiation is used.
The learned primary variables are normalized to a unit range around zero, and the scaling matrix is explicitly in one step. The scaling matrix could be a way to systematically nondimensionalize the entire problem.

>>>>

The system of equations we are looking at are, for only two components,
\begin{align}
  \partial_t \rho_{H_2O} =& \nabla \cdot F_{H_2O} + s \\
  \partial_t \rho_{CO_2} =& \nabla \cdot F_{CO_2} + s \\
  \partial_t e =& \nabla \cdot F_{heat} + s
\end{align}
More species add more balance laws (rows of equations). The applications also involve continuum
mechanics and chemical reactions, which are not considered. The interplay of deformations,
reactions, and phase transitions makes the equations extremely difficult
to solve. \cite{queiruga2018millstone}
(In this paper, we use $\partial_t$, $\mathrm{d}/\mathrm{d}t$,
and $\dot{()}$ to denote differentiation with respect to time
interchangeably depending on context.
The complexity is hidden inside the relationships between temperature
$T$, and pressure,$p$, density, $\rho$, and internal energy $e$ (or
enthalpy, $h$, s.t. $e=\rho h - p$), 
for the materials in question.

% {\bf RESERVOIR PHYSICS LIT REVIEW HERE; Other applications;
% plasticity, }
One of the simulators with the widest-encompassing physical modules is
the TOUGH family of solvers that have been developed at Lawrence
Berkeley National Lab.
The methodology is based on a primary variable switching method in 
The simulator of THAT GUY WITH DUAL VARIABLES I MET AT AGU
parameterizes a simplified hydrate system using density and enthalpy
as the primary variables.
Higher order representations of flow are also a problem to tackle
anisotropy and increase the computational efficiency of the spatial
discretization. CITE ME AND MICKEY AND STEPHAN
Multiphase flow was done with finite elements in \cite{YANG2014417}.
THAT MORIDIS FEM PAPER
However, the complexity of the state representation restricts
development to a 2-phase

@@

We redescribed our problem as a differential algebraic equation
with a constraint: the balances of mass and energy are constrained by
the EOS. We can cast the above equation into a basic form for a
differential algebraic equation,
\begin{align}
  z( u, \partial_t u) &= 0 \\
  c(u) & = 0
\end{align}
where, for the equation above, the variable is $u=\{T,p,\rho,h\}$. We will introduce a new way of thinking about constraints as sets of
solutions or observations of possible $u$:
\begin{equation}
c(u)=0 \longleftrightarrow \left\{ u | c(u)=0 \;\text{or}\; u\; \text{observed} \right\}
\end{equation}
This maps directly to the original experimental procedure for
determining a constraint equation.

@@

 A program synthesis approach was used to search for and evaluate possible methods of Peridynamics in Queiruga, 2017 \cite{queiruga2017numerical}. (An open source implementation of the study can be found Periflakes\cite{queiruga2018periflakes}.)

@@

We instead replace our problem with two balance laws and one unknown
constraint:
\begin{align}
\text{Solve for}\, \rho(t), \, h(t), \, p(t),\, \text{and}\, T(t)\, \text{satisfying:}\\
\partial_t \rho &  = \mathbf{k}_p(p_\infty - p) + r\\
\partial_t (\rho h-p) & = \mathbf{k}_T(T_\infty-T)  + s + h\left(\mathbf{k}_p(p_\infty - p) + r\right) \\
\text{such that they lie on the material EOS,}\\
\rho,h,p,T & \in \{\text{EOS}\}
\end{align}
The \emph{ground truth} is the experimental data in the first place; the
branching curve fit is only one realization of representing the data. We
have demonstrated that it is not possible to represent any two of these
in terms of the other two; and even if so, it is not necessarily a good
and robust fit. What if we search for a new fit?

@@

\begin{equation}
\bar{H}_{ij} = H_{ij} + \left(\delta_{ij}\,\text{if}\,H_{ik}=0\forall k\right)
  \end{equation}

@@

The architecture is illustrated in Figure \ref{fig:classifyingnetwork}.
It has the following form in index notation:
\begin{align}
  q_i &= W^{enc}_{ij} x^n_j + b^{enc}_i\\
  f_{ij} &= W^{dec}_{ijk} q^m_k + b^{dec}_{ij} \\
  h_i &= \mu\left(W^{bound}_{ik}q^m_k+b^{bound}_{i}\right) \\
  p_i &= \sigma\left(W^{phase}_{ij}h_j+b^{phase}_i\right) \\
  \bar{x}_i&= f_{ij} p_j
\end{align}
where $x$ is the original prediction, $x^n$ denotes a polynomial basis
set on $x$ (which increases the number of channels), $q$ is the latent
space, $f$ is a set of polynomial curves, $h$ is a hidden layer
denoting sides of boundaries, $p$ is a classification layer depending
on $h$, and $\bar{x}$ is the predicted feature. The activation $\mu$
is chosen as a standard neuron, e.g. $\tanh$ or $\text{sigmoid}$. The
activation $\sigma$ is a probability activation, which is chosen to be
a softmax,
\begin{equation}
\sigma_i(x;\beta) = \frac{e^{\beta x_i}}{\sum_j e^{\beta x_j}}
\end{equation}
or, in the limit, the hard argmax:
\begin{equation}
\sigma_i \rightarrow \text{onehot}\arg\max z as  \beta\rightarrow\infty
\end{equation}

@@

\begin{figure}
  \centering
  \includegraphics[width=3in]{/Users/afq/Documents/Dropbox/ML/primaryautoencoder/slides/classifier_good.jpg}
  \caption{\label{fig:classified}Result of the unsupervised phase classifier. Compare this plot to the original labeled dataset Figure \ref{fig:pTrho}.}
\end{figure}
Because the identity of the phases,XXXXXXXX
For comparison purposes, we group together liquid-gas-supercritical as one contiguous label.

@@

Examining the $T,p,\rho$ surface allows for human interpretability, but
the decoder actually yields four new functions with two unknowns:
$D(q)=\{T(q),p(q),\rho(q),h(q)\}$. These surfaces can be plotted as
independent 2D functions: the surface of \ref{fig:classified} is
unfolded into the new parameterization in Figure \ref{fig:foursurfaces}.
\begin{figure}
  \centering
  \includegraphics{../slides/water_four_surfaces_plot.png}
  \caption{\label{fig:foursurfaces}The four 2D surfaces that make up
    the new parameterization. The learned surface captures the sharp
    kinks and vertical fronts in Figure \ref{fig:classifyingresult},
    but is an analytic and smooth function of the two learned unknowns.}
  \end{figure}

With the trained model, we solve for \(q_1(t)\) and \(q_2(t)\) such that:
\begin{align}
\partial_t \rho(q_1,q_2) & = \mathbf{k} \left( p_\infty-p(q_1,q_2) \right)  + r \\
\partial_t \left(\rho h(q_1,q_2)-p\right) & =\mathbf{k}_T(T_\infty-T(q_1,q_2))  + s + h\left(\mathbf{k}_p(p_\infty - p) + r\right)
\end{align}
where the expressions for  \(\rho(q_1,q_2)\) etc. are the back of the
trained autoencoder,. The trained
autoencoder automatically removes the local constraint, satisfying
\begin{equation}
eos(D(q_1,q_2)) = 0 \, \forall \,q_1,q_2
\end{equation}
Thus, the system will always satisfy the constraints as $q$ is
evolved, even during intermediate iterative solution steps. Note again
that the machine learning is only applied to the
material representation: the balance laws are expressed directly.

To develop the rest of the discretization, we introduce the compact form:
\begin{align}
\frac{\mathrm{d}}{\mathrm{d}t} m(T,p,\rho,h) &= r(T,p,\rho,h)\\
eos(T,p,\rho,h) &= 0
\end{align}
where $m=\{\rho,\rho h -p\}$ and $r$ is the right hand side, with no good representation of \(eos\) everywhere.

The decoder part of the network is inserted into the balance law
\begin{equation}
\frac{\mathrm{d}}{\mathrm{d}t} m(D(q)) = r(D(q))
\end{equation}
We can parameterize this equation with any time stepping scheme, .e.g. a
Runge-Kutta scheme. For subsurface flow, backward Euler is popular due
to its simplicity and L-stability, which we follow since the topic of
this paper is not exploring new time stepping schemes. This yields the
discretizated algebraic equation for \(q_i\) and \(t+\Delta t\) given the
value \(q^0\) at \(t\),

@@

This equation is deceptively simple: the function $R(q)$ has complex logic encoded in it; if-then
statements appear if the model used activations such as rectifiers.
However, the complex material behavior which traditionally requires
the vast majority of programming was implemented automatically by the
models; implementing equations XX, XX, and XX requires only a few lines
of human-written code.

\subsection{Dimensionalization}

Recall that the datasets are normalized to be zero-center and
unit-ranged to improve training. Thus, the outputs of the decoder are not in true
units.
The ranges are saved in a separate file which are loaded by the
simulation. The network is loaded, the $D(q)$ is extracted, and the following
transformations are applied to the outputs:
\begin{align}
T(q)      &= \left(T_{max}-T_{min}\right) D_T(q) +T_{mean} \\
p(q)      &= \exp\left(\left( p_{max}-p_{min}\right)
            D_p(q) +p_{mean} \right) \\
\rho(q) &= \left(\rho_{max}-\rho_{min}\right) D_{\rho}(q) +\rho_{mean}\\
h(q)      &= \left( h_{max}-h_{min}\right) D_h(q)  +h_{mean}
\end{align}
The exponent transformation is only applied to the cases where
$\log(p)$ was used to make the training set.
The range values are in the proper units, so that equation rows in the
simulation are in SI units.

@@

There is a caveat with representing the degrees of freedom of the system
on a learned space: the user does not know the meaning of $q$ as an
input. Further, the values of $q$ vary from network to network and
depend on the parameterization. The user needs to be able to specify the
"real" physical quantities.

Since the original manifold is 2D, the user only needs to specify two
initial values. In a multiphase solver such as TOUGH, the state needs
to be specified with the two primary variables plus the state tag (sans input file formatting): 
\begin{verbatim}
    9e5 20 Aqu
    0.4 20 AqG
\end{verbatim}
Note that the first primary variable means pressure in the first line
for a pure-liquid state,
but saturation in the second line for the liquid-gas equilibrium.
This makes sense to human intuition but
still leads to bugs and occasionally requires looking up values in
textbooks or the software manual. (Is it $S_{gas}$ or $S_{aqu}$ on the second
line?) A fully fledged solver needs to check the user specified
initial conditions for erroneous inputs, but it cannot detect ambiguities.

The extra process of determining the initial state means that more error checking is needed
in verifying user inputs, but this is necessary in any
multiphase/multicomponent reservoir simulator. LatentSim does not yet
detect and report errors, and only reports divergence. In our study, the
testing problem specifications are all correct, but not all network
realizations are acceptable. Failure to determine a \(q\) for one of the
test initial conditions is a common mode of detecting unacceptable
network realizations; this is marked as a testing error during training
which invalidates a model as a potential realization.

@@

Note that the encoder only necessarily works when all four values are
correct. Thus, it is necessary to actually solve the nonlinear problem
on the decoder:
\begin{equation}
  D(q^0)_i = s^0_i \quad \text{for specified}\; i
\end{equation}
This equation has two rows indexing into the four state variables;
e.g. only the rows associated with $p$ and $T$ are solved if those are
specified, which is usually the case. The above equation is solved
with Newton's method. The two corresponding rows are cut out of
the tangent matrix to $D$,
\begin{equation}
K_{ij} = \frac{\partial D_i}{\partial q_j}  \quad \text{for
    specified}\; i, \text{for}\;j=1,2
\end{equation}

and the following system of equations is solved for an update to $q$,
\begin{equation}
K_{ij}\Delta q_j = s^0_i-D(q)_i
\end{equation}
The iteration on $q$ is repeated until convergence, with under-relaxation used again to update:
 \begin{equation}
q^0_{k+1} =q^0_{k} + \alpha \Delta q_{k}
\end{equation}
with the under-relaxation parameter usually set to $\alpha \approx
0.1$.

The initial condition problem is not guaranteed to converge, however,
and often does not.
For example, if the initial guess was in the liquid phase and the target point
in the EOS space is in a solid, the iteration will fail.
To address this issue, we perform multiple starts with an initial condition in each
phase and equilibrium, discard points that diverge, and then pick the
\(q\) that is closest to the two specified components. The multiple
start points are specified at EOS development time to form a set of
values that represent middle points for each phase,
\begin{equation}
\left\{s_{guess}\right\} = \left\{ \left(\begin{array}{c}
\bar{T}_{gas}\\
\bar{p}_{gas}\\
\bar{\rho}_{gas}\\
\bar{h}_{gas}
\end{array}\right),\left(\begin{array}{c}
\bar{T}_{liq}\\
\bar{p}_{liq}\\
\bar{\rho}_{liq}\\
\bar{h}_{liq}
\end{array}\right),\left(\begin{array}{c}
\bar{T}_{ice}\\
\bar{p}_{ice}\\
\bar{\rho}_{ice}\\
\bar{h}_{ice}
\end{array}\right)... \right\}
  \end{equation}
The multistart method could be specified without any prior knowledge
with random points, but this is a sufficiently robust method for
now. Each of the possible guess points is used to fill in the two
unspecified values as guesses and then passed into the encoder.
E.g. if $p$ and $T$ were specified, each possible $\bar{\rho}$ and
$\bar{h}$ combination from the above sets are filled in to provide
multiple guesses for $q$,
\begin{equation}
s^0 = \left(\begin{array}{c}
T^*(0)\\
p^*(0)\\
\bar{\rho}_{guess}\\
\bar{h}_{guess}
\end{array}\right);\quad q^0_{guess}=E\left(s^0\right)
 \end{equation}
 These sets of $q^0_{guess}$ are then iterated upon using the Newton's
 method procedure.

It is further possible that none of the initial guesses for this iteration
converge. There are two possibilities: the target state point is
incorrectly specified or out-of-range (e.g., negative density), or the
network is incapable of representing the desired material properties
(i.e., poorly trained). The first possibility shows an invalid
problem specification for the chosen EOS. The second possibility is
crucial to the automated training process. If a given network cannot
find an initial condition for every possible test problem, it fails
the tests and is marked as invalid. This is one way of automatically
verifying the training process.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Make a database of \(T,p,\rho,h,X_1,...\)
  \begin{itemize}
  \item
    Piece together empirical fits for each phase from literature
  \item
    (\emph{Experimental data in the future})
  \end{itemize}
\item
  Normalize (and \(\log(p)\)) the database, shuffle it

@@

The codebase is released open source at
\url{https://github.com/afqueiruga/latentsim}. Helper methods and
custom optimizers that the
author reuses in various TensorFlow-based projects are factored out in
the dully-named \url{https://github.com/afqueiruga/afqstensorflowutils}.
The implementation of the empirical equations of state for water are
factored out at
\href{https://github.com/afqueiruga/equations_of_state}{https://github.com/afqueiruga/equations\_of\_state}.

@@

The phase space of water is broken up into three different extents to
change the complexity of the training process:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item Linear Equation of State: the range is $ p = 10^5+[-10^3, 10^3]
  Pa,\quad T = [ 19, 21 ] ^o C$. This reduces to the single phase
  Darcy's law problem where the material has constant compressibility
  and heat capacity. This is the most basic test of the training
  framework, and provides analytical solutions to test the simulation
  framework.
\item Liquid-Gas Regime: the range is  $ p = [100,5\times 10^5] Pa,
  \quad T = [274,594] K$. This restricts the domain to only one phase
  boundary. With this range, it is possible to find a linear
  combination of $T,p,\rho,h$ that can make a 1-1 mapping. (The usual
  $T-p$ combination does not, but $\rho-h$ would.)
\item Solid-Liquid-Gas-Supercritical Regimes: this covers the full
  extent of the parameterized equation of state of water,
    $ p = [6\times 10^{-6},3\times 10^8]Pa, \quad T = [150,1000] K
    $. This problem has no linear mapping to a unique latent space.
\end{enumerate}

@@

The verification problems are listed in Table \ref{tab:problems}.
\begin{table}
  \caption{\label{tag:problems}Enumeration of the problems used to test and evaluate each fo the generated architectures}
\begin{tabular}{l|c|c|c}
  Name & start & path & end\\
\hline
\end{tabular}
\end{table}
The simulator was benchmarke against the long standing TOUGH+ multiphase
multicomponent

Single-grid block simulations were set up for most of the cases. TOUGH+
does not handle supercritical \ch{H2O}, so these test problems were not
used to benchmark. (The supercritical phase region is not practically
relevant for water, but it is very important for other fluids such as
\ch{CO2}.) (Note that, as a major motivation of this work, adding
supercritical states to the existing Fortran simulator would require
coding of one new logical state with two new state transitions with
primary variable switches; for the approach of this paper, the dataset
just needed to be extended.)

@@

We illustrated a programming model that is able to incorporate machine
learning models into existing well known computational methodologies.
This allows us to enforce physically known quantities, such as the
balance laws.

The TensorFlow system is not quite suited for this type of workflow -\/-
the resulting simulations are extremely slow due to repeated calls to
session.run and a very complicated graph structure for the polynomials.
Currently, the system is being reimplemented in Julia using Flux and
Zygote to seamlessly reincorporate into a fully-fledged new reservoir
simulator. The pendulum has been reimplemented in Julia at .

@@

Each implemented equation of state (three options in this work spanning
different extents of \ch{H2O}) has a database of potential realizations.
The entries are model architectures, parameters, and training
environment configurations and logs for each equation-of-state dataset.
This database of models replaces a version controlled repository of user
written modules. Hundreds of trials for novel primary variable
configurations happen in a matter of minutes on a workstation in place
of pain-staking trial-and-error labor performed with human "intuition"
and experience. We did encode a little bit of human intuition into the
model architectures and tricks for initializing the networks.

@@

In TOUGH, for one instance, all of the calculations are performed in SI units because it is simply too much labor to normalize everything! Adverse effects to iterative algorithms and floating point error propogation occur, especially because numerical differentiation is used.
The learned primary variables are normalized to a unit range around zero, and the scaling matrix is explicitly in one step. The scaling matrix could be a way to systematically non-dimensionalize the entire problem.
